Compositional Quantitative Resource Analysis of C Programs

### 2014-06-05 paper/main.tex
cleanup in paper

\section{Introduction}

After functional correctness, quantitative properties are the
second thing one wants to know when writing a program.
For example, an asymptotic runtime bound on a
sorting algorithm tells us what efficiency we can expect
of it when running on very large arrays.  But, what if we
always run it on arrays of size between 5 and 10? In that
case, the asymptotic is not relevant and we need a more
precise bound on the runtime.  It is then a good
software engineering practice to profile the algorithm on
some examples.

In between asymptotic bounds and profiling there are
precise bounds, for instance, one can prove that Knuth's Quicksort might
take at most $20N^2$ operations of a MIX machine when
called on an array of size $N$.  If you have this information,
no profiling is needed to compare the algorithms and you
have a formal guarantee on the behavior of the program. But what if you do not
have a MIX machine, or you tweaked Knuth's algorithm
a bit?  Then you are good for another long an tedious
proof.  We advocate that, it is possible to design a practical
system that is able to infer precise worst case bounds
without any user interaction.

This work presents a resource oblivious system to get
linear resource bounds on imperative programs with
integers.  Despite its apparent simplicity it is able to
reproduce multiple results from the literature which
were obtained using methods based on abstract
interpretation.  In stark contrast with abstract-interpretation
based methods, our tool does not require any
fixpoint computations on an abstract domain to obtain
loop invariants.  The mechanism we designed, which
is based on amortized analysis, is able to leverage a local
information to obtain global invariants.  Additionally, the
approach we advocate is able to scale to full blown
imperative languages including constructs like functions
calls, \code{break} and \code{return}.

The analysis is proved sound with respect to
the operational semantics instrumented to track resource
consumption.  It is implemented as a prototype
in OCaml and available on demand.  We test it on multiple
algorithms from the literature and on custom tests, it
is also planned to extend it to handle the full C programming
language.

\section{Overview and Examples}

\subsection{Amortized Analysis}
The idea that drives the design of our system is amortized
analysis.  Say that a program $p$ evaluates on a heap $H$
by consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.
The basic idea of amortized analysis is to define a potential function
$\Phi$ from heaps to natural numbers, such that $\Phi(H) \ge n$.

>>>>

\section{Amortized Analysis for C Programs}
\label{sec:inform}

\subsection{Amortized Analysis}
The idea that drives the design of our system is amortized
analysis [XXX].  Say that a program $p$ evaluates on a heap $H$
by consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.
The basic idea of amortized analysis is to define a potential function
$\Phi$ from heaps to natural numbers, such that $\Phi(H) \ge n$.
